# Literature Review

```{r packages, include = FALSE}
library(tidyverse)
library(kableExtra)
```

Uncertainty has been examined in various ways over the last two decades,
and is becoming increasingly important for researchers. This review
looks at why uncertainty is important to evaluate in transportation
demand models, and research that has been done to evaluate uncertainty.
@rasouli2012 has an extensive literature review on this topic. An
overview of the literature and which source of uncertainty they evaluate
can be found in @tbl-authors.

```{r authors, echo = FALSE, results = 'asis', warning = FALSE}
#| label: tbl-authors
#| tbl-cap: Studies of Forecasting Uncertainty
authors_source <- tibble("Reference" = c(
  "@rodier2002uncertain",  
  '@zhao2002',  
  '@clay2005univariate',  
  '@flyvbjerg2005',  
  '@armoogum2009',  
  '@duthie2010highway',  
  '@welde2011planners',  
  '@yang2013',  
  '@manzo2015',  
  '@petrik2016measuring',  
  '@petrik2020uncertainty', 
  '@hoque2021'), 
  "Uncertainty Source(s) Evaluated" = c(
    "Input Data",  
    "Input Data & Parameter Estimates",  
    "Input Data & Parameter Estimates", 
    "Model Form", 
    "Model Form", 
    "Input Data & Parameter Estimates", 
    "Model Form", 
    "Input Data & Parameter Estimates", 
    "Input Data & Parameter Estimates", 
    "Input Data & Parameter Estimates", 
    "Model Form & Parameter Estimates", 
    "Input Data"))  
kbl(authors_source, 
    booktabs = TRUE, format="markdown", longtable = FALSE) |>  
  kable_styling()
```

Model accuracy is the basis for why uncertainty of input data and/or
parameter estimates are important to study. Travel forecasters have
always been cognizant of the uncertainty in their forecasts, especially
as project decisions are made using these models, often with high
financial impacts. 

@flyvbjerg2005 collected data from various forecasting traffic models
with an emphasis on rail projects. They used the forecast data for a
given year and the actual value that was collected for the same year.
Their study found that there is a statistical significance in the
difference of the estimated and actual values. Rail projects are
generally overestimating passenger forecasts by 106%, and half of road
projects have a traffic forecast difference of plus or minus 20%. They
did not identify where this inaccuracy came from, but they identified
that it was important for future research.

@armoogum2009 looked at uncertainty within a forecasting model for the
Paris and Montreal metropolitan regions. The sources of uncertainty
analysed were calibration of the model, behavior of future generations,
and demographic projections. A jackknife technique, rather than sampling
methods, was used to estimated confidence intervals for each source of
error using multiple years of analysis. This technique is a way to
reduce the bias of an estimator and permits the estimation of confidence
intervals to produce variance estimates. They found that the longer the
forecasting period was, the larger the uncertainty. Generally the model
forecast within 10-15%, reaching higher percentage ranges for variables
with small values or small sample sizes.

@welde2011planners compared actual and forecast traffic values for 25
toll and 25 toll free roads in Norway. They evaluated the accuracy of
Norwegian transportation planning models over the years. Generally
traffic models overestimate traffic. This study found that toll
projects, on average, overestimated traffic, but only by an average of
2.5%. Toll free projects, however, underestimated traffic by an average
of 19%. They concluded that Norwegian toll projects have been fairly
accurate, with a probable cause coming from the scrutiny that planners
get when developing a toll project. A similar scrutiny should then also
be placed on toll free projects as they are significantly less accurate.

These articles show that models have errors which effects traffic
projections by a significant amount. These articles identified that
error existed but did not quantitatively identify the source of the
error. The most researched error source has been on model form but that
research has mostly been excluded in this review as it is not the main
focus of this research. The second most researched form has been on
input data. Chronologically, @rodier2002uncertain, @zhao2002,
@clay2005univariate, @duthie2010highway, @yang2013, @manzo2015, and
@petrik2016measuring have all researched input error, with all but the
first also looking at parameter estimate error as well. Parameter
estimation error has been the least researched source of uncertainty,
where there have been no studies focused only on that source of error.
@petrik2020uncertainty looked at parameter estimates, but with a focus
also on model form error. The details of each study are described below
in chronological order.

@rodier2002uncertain looked at uncertainty in socioeconomic projections
(population and employment, household income, and petroleum prices) at
the county-level for the Sacramento, California region. They wanted to
know if the uncertainty in the range of plausible socioeconomic values
was a significant source of error in the projection of future travel
patterns and vehicle emissions. They identified ranges for population
and employment, household income, and petroleum price for two scenario
years (2005 and 2015). The ranges varied based on the scenario year and
the socioeconomic variable. They changed one variable at a time for a
total of 19 iterations of the model run for 2005 and 21 iterations for
2015. Their results indicated that the error in projections for
household income and petroleum prices is not a significant source of
uncertainty, but error ranges for population and employment projections
are a significant source for changes in travel and emissions. The input
data of population and employment were a significant factor to the model
result uncertainty.

@zhao2002 looked at the propagation of uncertainty through each step of
a trip-based travel model from variation among inputs and parameters.
This analysis used a traditional four-step urban transportation planning
process (trip generation, trip attraction, mode split, and trip
assignment) on a 25-zone sub-model of the Dallas-Fort Worth metropolitan
region. Monte Carlo simulation was used to vary the input and parameter
values. These values were all ranged using a coefficient of variation
($c_v$) of 0.30. The four-step model was run 100 times with 100
different sets of input and parameter values. The results of these runs
showed that uncertainty increased in the first three steps of the model
and the final assignment step reduced the compounded uncertainty,
although not below the levels of input uncertainty. The authors
determined that uncertainty propagation was significant from changes in
inputs and parameters, but the final step nearly stabilizes the
uncertainty to the same amount as assumed (0.30 $c_v$ assumption with a
0.31 $c_v$ in the results of trip assignment).

Another study that looked at input data uncertainty was
@clay2005univariate. These researchers varied three inputs and one
parameter to analyze uncertainty of outputs on a fully integrated land
use and travel demand model of six counties in the Sacramento,
California region. The variables used for analysis were productions,
commercial trip generation rates, perceived out-of-pocket costs of
travel for single occupant vehicles, and concentration parameter.
Exogenous production, commercial trip generation rates, and the
concentration parameter were varied by plus or minus 10, 25 and 50%,
while the cash cost of driving was varied by plus or minus 50 and 100%.
This resulted in 23 model runs, one for each changed variable and one
for the base scenario. Their research found that any uncertainty in the
inputs resulted in large difference in the vehicle miles traveled
output, although this difference was a lower percentage than the
uncertainty in the input.

@duthie2010highway evaluated uncertainty at a different level. They use
a small generic gravity-based land use model with the traditional four
steps, using a coefficient of variation of 0.3 from @zhao2002 for input
and parameters, although using antithetic sampling. In this sampling
method, pairs of negatively correlated realizations of the uncertain
parameters are used to obtain an estimate of the expected value of the
function. The uncertainty was evaluated on the rankings of various
transportation improvement projects. They found that there are a few
significant differences that arise when changing the input and parameter
values that result in different project rankings, and thus neglecting
uncertainty can lead to suboptimal network improvement decisions.

@yang2013 evaluated a quantitative uncertainty analysis of a combined
travel demand model. They looked at input and parameter uncertainty
*also* using a coefficient of variation of 0.30. Rather than using a
random sampling method for choices they used a systematic framework with
a variance-covariance matrix. Their research found that the coefficient
of variation of the outputs are similar to the coefficient of variation
of the inputs, and that the effect of parameter uncertainty on output
uncertainty is generally higher than that of input uncertainty. This
finding contradicts the finding of @zhao2002. The authors concluded that
improving the accuracy of parameter estimation is more effective that
that of improving input estimation as they found that in most steps of
the model, the impact of parameter uncertainty was more important that
that of input uncertainty.

@manzo2015 looked at uncertainty on model input and parameters for a
trip-based transportation demand model in a small Danish town. They used
a triangular distribution with LHS to create the range in parameters,
and using the information from @zhao2002 they also used a coefficient of
variation of 0.30 and 100 draws, choosing these values at they had been
previously used. Their addition to the research of uncertainty, was by
examining uncertainty under different levels of congestion. Their
research found that there is an impact on the model output from the
change in input and parameter uncertainty and requires attention when
planning. Also, model output uncertainty was not sensitive to the level
of congestion.

@petrik2016measuring evaluated uncertainty in mode shift predictions due
to uncertainty from input parameters, socioeconomic data, and
alternative specific constants. This study was based on a high-speed
rail project in Portugal as a component of the Trans-European Transport
Network. They collected survey data and developed discrete choice
models. The authors created their own parameter values from the
collected data, obtaining the mean or "best" value from the surveys and
the corresponding t-statistic. With these they generated 10,000 samples
each of parameter values, socioeconomic inputs, and mode-specific
constants, using bootstrap re-sampling, Monte Carlo sampling, and
triangular distribution methods respectively. The authors found that
variance in alternative specific attributes is the major contributor to
output uncertainty in comparison to parameter variance or socioeconomic
variance. Socioeconomic data had the least contribution to overall
output variance, and there was a relatively insignificant mode shift due
to variability in parameters.

@petrik2020uncertainty used an activity based microsimulation travel
demand model for Singapore to evaluate model form and parameter
uncertainty. This model has 22 sub-models and 817 parameters. The
authors determined which of the 817 parameters the sub-models were most
sensitive to and applied a full sensitivity analysis of the top 100 of
the parameters, preserving correlations. Using the mean parameter value
and the standard deviations they had for all of them they used Latin
hypercube sampling with 100 draws to look at the outcomes of the change
in each parameter value. Different sized samples of the model population
were also considered in their research. They found that of the 100 most
sensitive parameter values, the outcome coefficient of variation varied
from 3% to 49%. The variance of the parameter variables did not exceed
19%, and thus the results from the parameter uncertainty were higher
than the variance in the parameters. They also found that the results of
the parameter uncertainty was higher than simulation uncertainty.

In transportation demand models, when uncertainty is analysed, most
error research to this point has focused on input uncertainty or model
forms, rather than parameter estimate uncertainty [@rasouli2012]. In
this literature review 12 articles have been included. Of those 12, two
look at input data as the only focus of their uncertainty research,
three focus on model form uncertainty, one looks at both model form and
parameter estimate uncertainty, and six focus on both input data and
parameter estimate uncertainty. No researchers have looked at parameter
estimate uncertainty as the only source of error in their models. This
is a gap in uncertainty research and is why that for this study, the
parameter estimates within the model uncertainty is of the most
immediate concern. When parameter uncertainty has been examined in
literature it is often in conjunction with input errors, or on
relatively small and impractical models. No researchers have used real
models for their analyses. There have also been some conflicting
conclusions on how parameter variance impacts output uncertainty. There
is a research need within uncertainty analyses to have additional and
repetitive studies that can confirm the outcome of previous research,
since it is a relatively new topic of discussion. Uncertainty research
is needed as transportation demand models provide estimates and
forecasts for decision and policy makers. An inaccurate model or large
output variance could change what decisions are made and when
[@aep50_2023]. There is also a research need within uncertainty for both
activity-based models, and destination and transport mode choice model
components. In this research, I investigated the variance in forecast
traffic volumes resulting from varying the mode and destination choice
parameters in an advanced trip-based travel demand model, using combined
mode and destination choice models, coupled with highway assignment.
